\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms

### 2013-05-14 paper/paper.tex
added to paper

% This version uses the latex2e styles, not the very ancient 2.09 stuff.
\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{epsfig,graphicx,usenix}

@@

%don't want date printed
\date{May 2013}

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms}
%for single author (just remove % characters)
\author{
{\rm Colleen Josephson\footnote{Supported in part by Lindt Chocolatier.}}\\
MIT
\and
{\rm Pavel Panchekha}\\
MIT
% copy the following lines to add more authors
% \and
% {\rm Name}\\
%Name Institution
} % end author

@@

@@

Sample all the rates~\cite{samplerate, sprout, minstrel}.

@@

Minstrel utilizes the multi-rate retry chain (mrr), which is an array that tells the card which rates to try before reporting a failure. The retry chain makes failures extremely unlikely. Minstrel has a mrr of [highest throughput, next highest throughput, best probability, lowest base rate] for most packets. In essence, Minstrel tells the card to send at the highest throughput rates but to fall back to reliable rates upon failure. The throughput and success probability are calculated for each rate every 100ms. 
To keep an accurate estimate of throughput and probability for each rate, Minstrel sends sample frames ten percent of the time. If the randomly chosen rate has a higher lossless throughput than the current optimal rate, the mrr is as follows: [random, highest throughput, best probability, lowest base rate]. If the random rate is slower than the optimal rate, it is moved further back in the retry chain: [best throughput, random, best probability, lowest base rate]. This ensures that Minstrel never samples rates worse than the current optimal rate unless the optimal rate experiences a failure.

@@

Yeee haw.
\begin{figure}[htb]
  \hspace{-4em}\includegraphics[width=3.75in]{constant.png}\vspace{-2em}
  \caption{dfg}
\end{figure}

@@

@@

\subsection{Availability}
We should make our github repo or something like that available here.

@@

802.11n, SampleRate with EWMA, take data using broadcast approach,  submitting minstrel changes as kernel patch?

@@

\section{Acknowledgements}
Jonathan Perry, Hari Balikrishnan, and Derek Smithies.

>>>>

% This version uses the latex2e styles, not the very ancient 2.09 stuff.
\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{epsfig,graphicx,usenix,fullpage, hyperref}

@@

%don't want date printed

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms}
%for single author (just remove % characters)
\author{
{\rm Colleen Josephson}\\
cjoseph@mit.edu
\and
{\rm Pavel Panchekha}\\
pavpan@mit.edu
% copy the following lines to add more authors
% \and
% {\rm Name}\\
%Name Institution
} % end author

@@

SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bitrate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.

@@

@@

Minstrel utilizes the multi-rate retry chain (mrr), which is an array that tells the card which rates to try before reporting a failure. The retry chain makes failures extremely unlikely. Minstrel has a mrr of [\emph{highest throughput, next highest throughput, best probability, lowest base rate}] for most packets. In essence, Minstrel tells the card to send at the highest throughput rates but to fall back to reliable rates upon failure. The throughput and success probability are calculated for each rate every 100ms. 
To keep an accurate estimate of throughput and probability for each rate, Minstrel sends sample frames ten percent of the time. If the randomly chosen rate has a higher lossless throughput than the current optimal rate, the mrr is as follows: [\emph{random, highest throughput, best probability, lowest base rate}]. If the random rate is slower than the optimal rate, it is moved further back in the retry chain: [\emph{best throughput, random, best probability, lowest base rate}]. This ensures that Minstrel never samples rates worse than the current optimal rate unless the optimal rate experiences a failure. See table ~\ref{table:1} for an overview of the mrr's used in Minsterl.

Minstrel's implementation is based on SampleRate's, so it also has \texttt{ApplyRate()} and \texttt{ProcessFeedback()}, as well as a \texttt{UpdateStats()} method that runs every 100ms. \texttt{UpdateStats()} is home to some of the key differences between Minstrel and SampleRate. Instead of making decisions based on the average 
transmission time, Minstrel uses throughput: \begin{equation} T = \frac{p}{tx\_time}\end{equation} where $T$ is the throughput of some rate $r$, $p$ is the probability of successfully transmitting at $r$, and $tx\_time$ is the lossless transmission time at $r$.

The probability $p$ of a successful transmission is calculated from statistics collected by the \texttt{ProcessFeedback()} method, and is done using an exponential weighted moving average, or EWMA. The EWMA creates a weighted average that weighs recent data more heavily than old data. The weighting for old data decreases exponentially. An EWMA is beneficial because it ensures that a sudden degredation in link quality will create a rapid response in the probabilities due to the increased weight of recent data. Minstrel will waste less time sending packets at a rate that no longer works, as compared to a simple probability calculation.

\begin{table*}[htb]
    \begin{tabular}[htb]{l|l|l|l}
    Try & Normal               & Random (slower than best) & Random (faster than best) \\ \hline
    1   & Best throughput      & Best throughput           & Random                    \\
    2   & Next best throughput & Random                    & Best throughput           \\
    3   & Highest probability  & Highest probability       & Highest probability       \\
    4   & Lowest rate          & Lowest rate               & Lowest rate               \\
    \end{tabular}
    \caption{Multi-rate retry chains for Minstrel}
\label{table:1}
\end{table*}

@@

\begin{figure}[hbt]
  \hspace{-0.5em}\includegraphics[width=3.5in]{bsmt13floorplan.png}\vspace{-0em}
  \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
\end{figure}

Yeee haw.
\begin{figure}[htb]
  \hspace{-.75em}\includegraphics[width=3.5in]{constant.png}\vspace{-0em}
  \caption{Throughput of using a constant 802.11b/g bitrate in an MIT basement.}
\end{figure}

@@

We implemented SampleRate as outlined in John Bicket's Master's thesis~\cite{samplerate}. This implementation is slightly different from how it was implemented in the MadWifi kernel driver. The primary difference is that the kernel implementatiion uses a EWMA instead of a 10 second window. The thesis implmenentation throws out old rate data and begins anew every 10 seconds. It would be worth looking at the performance of EWMA-based SampleRate in the future.

\texttt{samplerate.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{RemoveStaleResults()}, as well as a few helper functions and data structures for tracking rate statistics.

We implemented Minstrel by porting the C code from the 3.3.8 version of the Linux kernel into Python. \texttt{minstrel.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{UpdateStats()}, as well as a few helper functions and data structures for tracking rate statistics.

@@

\subsection{Availability}
All of our code and collected trace data are available on GitHub:

\noindent
\url{https://github.com/pavpanchekha/6.829-project/tree/3.8.6}

This repository contains the modified ath9k driver used to collect the traces, the traces we collected, as well as the Python simulation framework.

@@

Very little analysis has been done of the performance of bit rate selection algorithms on 802.11n networks. 802.11n introduces many new rates, as well as multiple-input/multiple output, or MIMO, capabilities. There is an 802.11n version of Minstrel, but there is no implmentation of SampleRate.

We implemented SampleRate as outlined in the thesis, but the MadWifi implementation has a few key differences from the thesis. We suspect that the MadWifi use of an EWMA instead of a window has more accurate rate statistics, which will potentially lead to better throughput. We did not have the time to implement this alternative version, but it is worth further examination.

Hari Balikrishnan suggested that we use a broadcast approach when collecting our traces, instead of link layer acknowledgements. This would entail having multiple computers listening at the broadcast address, and compiling successes and failures from that information. This approach would free us from certain idiosyncracies of reading data from the driver, and lead to possibly more accurate traces. We were unable to implement this due to a lack of time and ability to acquire compatible hardware.

Finally, we are considering submitting our improvements to the Minstrel algorithm, namely the BEWMA, as a kernel patch. Currently is implemented in Python, so we would have to port the code into C.

@@

\section{Acknowledgements}
We would like to thank Jonathan Perry and Hari Balikrishnan for their valuable guidance, as well as Derek Smithies for answering our questions about Minstrel.
